---
title: "Agentic AI for Credit Risk Assessment using LightGBM"
author:
  - name: "Supriya Sathe (Advisor: Dr. Cohen)"
    url: https://github.com/supriyasathe

format:
  html:
    mathjax: true
    theme: lightly

bibliography: references.bib
link-citations: true
---


## 1. Introduction

Evaluation of credit risk is vital for the financial services industry because, it is required to  avoid risks and maintain compliance with regulations, organizations have to accurately predict a borrower's likelihood of failure while re-payment. Because they are straightforward and easy to understand for anyone, typical evaluation methods like systems based on rules and regression analysis have been often used in AI world. Modern advancement in machine learning, although have shown that the hierarchical (structural) financial data can now be predicted with significant advantages.

The Risk of default Assessment, which measures a lender's probability count of defaulting lenders,  it is an important aspect of organizational stability in the technological current days in financial environment. The industry has already started moving from restrictive rule-wise structures to complicated Machine Learning (ML) architectures, as stated by Shreya and Pathak (2025) @shreya2025explainableartificialintelligencecredit. Although the fact is these models provide past unheard-of precision in forecasting, they are frequently addressed to as "black boxes." The paper uses Explainable Artificial Intelligence (XAI) to overcome the visibility gap and LightGBM to develop a robust credit assessment engine.

According to author de Lange et al. (2022) @de2022explainable and Bussmann et al. (2020) @bussmann2021explainable, financial businesses should offer a “Right to justification” as a way to make sure that risk management is the responsibility for both moral and verifiable. This flexibility is not just a technical preference in organization the but also a legal need. In a similar vein, de Lange et al. (2022) @de2022explainable demonstrate that explainable AI methods can maintain accurate predictions while working on to improve an  accessibility in financial systems nowadays. The significance of combining  the both concepts explainable machine learning with business analytics for ethical evaluation of credit risk defaulters have been demonstrated by recent studies by Shreya and Pathak (2025) @shreya2025explainableartificialintelligencecredit and Oyeyemi et al. (2025) @oyeyemiexplainable.

Here are the important steps to develop a machine learning-based credit risk model to build overall good performance:

1.	The first is collecting and preparing the data, including the encrypting variables with categories and finding the missing values.
2.	Dividing the dataset into two subsets for testing and training.
3.	Applying the training data to train for the LightGBM classifier.
4.	Assessing predictive performance for the the model with metrics like ROC-AUC, accuracy, precision, recall, and F1-score.
5.	Utilizing SHAP-based explanation and characteristic significance for analyzing the model forecasts.

This method provides that the model is easy to understand, versatile, and trustworthy. The predictive model's accuracy on unidentified information is determined using model evaluation approaches including cross-validity and train-test partitioning.

This project is using the Light Gradient Boosting Machine (LightGBM) programmed in Python langauge to create an Explainable Credit Risk Assessment system, driven by major breakthroughs in both research and business. LightGBM is determined because it performs well on structured table type datasets and is effective and versatile way. The Credit Risk Dataset's consumer demographic details, banking, and credit historical information will be used by the algorithm to forecast defaults listing on loans potential. Explainable methods like SHAP will be used to understand both specific and worldwide projections in order to assure integrity and transperency.

The objective of this study is to show how explainable artificial intelligence concepts may be integrated with efficient boosting algorithms to create precise, accessible, and insightful credit risk evaluation solutions to avoid the losses of financial organization.

---

## 2. Methods

### 2.1 Data Preprocessing
First, preprocessing is for the dataset that includes borrower demographic details, financial level, and credit history characteristics. To utilize statistical robustness and lessen vulnerability to extreme values, missing numerical data are handled via median imputation. One-hot encoding is used to convert categorical variables into numerical representation so that gradient boosting calculation can use them for prediction.

### 2.2 LightGBM Architecture for Credit Risk Prediction

This reasearch uses Light Gradient Boosting Machine (LightGBM) to predict the possibility of loan defaulter's based on borrower charactestics like, how much they are earning, employment history, how much load amount and history of credit. LightGBM is an really efficient as a boosting framework to make decision making tress sequentially to get accurate predictiona and reduce the errors in the prediction.

The above function is optimized by the LightGBM model:

$$
\text{Obj}(t)=
\sum_{i=1}^{n}
l\left(
y_i,
\hat{y}_i^{(t-1)} + f_t(x_i)
\right)
+\Omega(f_t)
$$

𝑦𝑖 indicates if the loan is in default or not. (default or non-default)

𝑦^i(t−1) is the value expected from the earlier loop.

𝑓𝑡(𝑥𝑖) shows the new decision tree that was added after iteration 

l(⋅) represent the binary cross-entropy value for classifier loss function.

Ω(ft) is the normalization idea that keeps excessive fitting from developing.

### 2.3 LightGBM Model Implementation 

### 2.4 Explainable Agentic AI Framework
An explainable agentic reasoning loop develop on the observe-reason-act model has been integrated with the system. The agent develops understandable explanations which is human-readable for credit risk evaluation through evaluating significant variables and observing predictive model outcomes through system.

The cycle of agentic reasoning is stated as:

Observation → Reasoning→ Decision → Explanation

To provide relevant risk assessment, the framework analyzes borrower some characteristics such as financial status, credit amount, background in employment, and credit history.

### 2.5 Model Training and Evaluation
This study spliting the dataset into training and testing subsets using stratified sampling and it keeps the balance of class classification. Gradient boosting iterations with optimized hyperparameters, such as learning rate and number of leaves, are used to train the model.

To guarantee dependable classification performance, model performance is assessed using ROC-AUC score, precision, recall, F1-score, and confusion matrix analysis.

## 3. Analysis and Results

### 3.1 Exploratory Data Analysis (EDA) : 

#### 3.1.1 Data Description :

The present research is using the Credit Risk Dataset, which contains 32,581 borrower records and twelve variables related to financial, credit history, and demographic details. A borrower's likelihood of defaulting (1) or not (0) is indicated by the binary target variable loan_status. The dataset includes both quantitative and qualitative factors, including income, loan amount, interest rate, length of work, home ownership status, and loan purpose. It is perfect for applying LightGBM and evaluating explainable, agentic AI decision-making in credit risk prediction due to its realistic structure and suitable size.



| Feature | Count | Unique | Top | Freq | Mean | Std Dev | Min | 25% | 50% | 75% | Max |
|----------|--------|--------|------|------|--------|---------|------|------|------|------|------|
| person_age | 32581 | – | – | – | 27.73 | 6.35 | 20 | 23 | 26 | 30 | 144 |
| person_income | 32581 | – | – | – | 66074.85 | 61983.12 | 4000 | 38500 | 55000 | 79200 | 6000000 |
| person_home_ownership | 32581 | 4 | RENT | 16446 | – | – | – | – | – | – | – |
| person_emp_length | 31686 | – | – | – | 4.79 | 4.14 | 0 | 2 | 4 | 7 | 123 |
| loan_intent | 32581 | 6 | EDUCATION | 6453 | – | – | – | – | – | – | – |
| loan_grade | 32581 | 7 | A | 10777 | – | – | – | – | – | – | – |
| loan_amnt | 32581 | – | – | – | 9589.37 | 6322.09 | 500 | 5000 | 8000 | 12200 | 35000 |
| loan_int_rate | 29465 | – | – | – | 11.01 | 3.24 | 5.42 | 7.90 | 10.99 | 13.47 | 23.22 |
| loan_status | 32581 | – | – | – | 0.22 | 0.41 | 0 | 0 | 0 | 0 | 1 |
| loan_percent_income | 32581 | – | – | – | 0.17 | 0.11 | 0.00 | 0.09 | 0.15 | 0.23 | 0.83 |
| cb_person_default_on_file | 32581 | 2 | N | 26836 | – | – | – | – | – | – | – |
| cb_person_cred_hist_length | 32581 | – | – | – | 5.80 | 4.06 | 2 | 3 | 4 | 8 | 30 |





**Dataset Dimensions** -

Observations (rows): 32,581

Features (columns): 12

Target Variable: loan_status

**Target Variable** -

loan_status

1 = High Risk (Default)

0 = Low Risk (Non-Default)

#### 3.1.2 Data Ingestion (Python) :

The pandas package in Python was used to load the dataset. A dataframe called credit1 was created by reading the file credit_risk_dataset.csv. The dataset was first explored by examining column data types, previewing the first few records, and determining its dimensions. To guarantee effective ingestion and comprehend variable composition, the dataset structure was validated using the.shape(),.head(), and.info() functions.

```{python}
#| code-fold: true
#| output: false

import os
import pandas as pd
from skimpy import skim

def find_git_root(start=None):
    if start is None:
        start = os.getcwd()
    path = os.path.abspath(start)
    
    while True:
        if os.path.isdir(os.path.join(path, ".git")):
            return path
        
        parent = os.path.dirname(path)
        if parent == path:
            raise FileNotFoundError("No .git directory found — are you inside a Git repository?")
        
        path = parent

repo_root = find_git_root()
datasets_path = os.path.join(repo_root, "datasets")

# Reading the datafile credit_risk_dataset.csv
credit_path = os.path.join(datasets_path, "credit_risk_dataset.csv")
credit1 = pd.read_csv(credit_path)
```

#### 3.1.3 Summary Statistics :

```{python}
import pandas as pd

pd.set_option('display.max_columns', None)
pd.set_option('display.width', None)

credit1.describe(include='all')
# credit1.columns

# for col in credit1.columns:
#     print(col)
```

For a more thorough summary:

```{python}
#| code-fold: true
from skimpy import skim
skim(credit1)
```

#### 3.1.4 Missing Value Analysis :

```{python}
credit1.isna().sum()
```

person_emp_length → 895 missing values

loan_int_rate → 3,116 missing values

Complete data is present in all other variables.

Preprocessing will use median imputation to handle these missing values in order to maintain the robustness of the distribution.


#### 3.1.5 Target Distribution :

Import the project dependencies (Libraries)

<!-- This code chunk is loading all libraries being used in the project. -->

```{r}
#| code-fold: true
#| message: false
#| warning: false
#| output: false

packages <- c("tidyverse", "skimr", "knitr", "ggthemes", "ggrepel", "dslabs")
# install.packages(packages)
# Load Libraries
lapply(packages, library, character.only = TRUE)
# Set seed for reproducibility
set.seed(123)
```

Here, I analyze target variable loan_status To show the distribution of default and non-default loan cases in the used dataset. I use value_counts() function to calculate the frequency counts and percentage percentages of each class using Python's Panda library. 
The Matplotlib Python's library uses to create a bar chart that depicted the count of borrowers who were in default and non-default list. This study analysis gives insight into the overall credit risk pattern and helps with appropriate model evaluation metrics and training decision-making.

```{python}
#| code-fold: true
#| echo: false
import matplotlib.pyplot as plt

# Target Distribution

target_counts = credit1['loan_status'].value_counts()
print("Target Counts:\n", target_counts)

target_percentage = credit1['loan_status'].value_counts(normalize=True) * 100
print("\nTarget Percentage:\n", target_percentage)

plt.figure()
credit1['loan_status'].value_counts().plot(kind='bar')
plt.title("Target Distribution (Loan Status)")
plt.xlabel("Loan Status (0 = Non-default, 1 = Default)")
plt.ylabel("Count")
plt.xticks(rotation=0)
plt.show()
```

```{python}
#| code-fold: true
#| echo: false
import matplotlib.pyplot as plt
import pandas as pd

# Load dataset inside Python environment
credit1 = pd.read_csv("datasets/credit_risk_dataset.csv")
plt.figure()
plt.hist(credit1['loan_amnt'], bins=30)
plt.title("Loan Amount Distribution")
plt.xlabel("Loan Amount")
plt.ylabel("Frequency")
plt.show()


```


```{python}
#| code-fold: true
#| echo: false
import pandas as pd
import matplotlib.pyplot as plt

# Load dataset inside Python chunk
credit1 = pd.read_csv("datasets/credit_risk_dataset.csv")

# Select numeric variables only
numeric_df = credit1.select_dtypes(include=['number'])

# Compute correlation matrix
corr_matrix = numeric_df.corr()

# Plot heatmap using matplotlib
plt.figure(figsize=(8,6))
plt.imshow(corr_matrix, interpolation='nearest')

plt.colorbar()
plt.title("Correlation Heatmap")

# Add axis labels
plt.xticks(range(len(corr_matrix.columns)), corr_matrix.columns, rotation=90)
plt.yticks(range(len(corr_matrix.columns)), corr_matrix.columns)

plt.tight_layout()
plt.show()
```



u### 3.1. Data Ingestion

This code chunk is importing the dataset from the folder datasets which it is finding at the root of the github repo and it is concatenating the path to run the function read_csv from tidyverse.

```{r}
#| code-fold: true
#| output: false
find_git_root <- function(start = getwd()) {
  path <- normalizePath(start, winslash = "/", mustWork = TRUE)
  while (path != dirname(path)) {
    if (dir.exists(file.path(path, ".git"))) return(path)
    path <- dirname(path)
  }
  stop("No .git directory found — are you inside a Git repository?")
}

repo_root <- find_git_root()
datasets_path <- file.path(repo_root, "datasets")

# Reading the datafile credit_risk_dataset.csv
credit_path <- file.path(datasets_path, "credit_risk_dataset.csv")
credit1 = read_csv(credit_path, show_col_types = FALSE)
```


```{r}
summary(credit1)
head(credit1)
# Count total NAs per column
colSums(is.na(credit1))
```

I found that:

- person_emp_length has 895 NA values
- loan_int_rate has 3116 NA values


test data ingestion using python

```{r}
# renv::use_python(type = "conda", envname = "myenv")
# reticulate::py_install("numpy")
# reticulate::py_install("pandas")
# reticulate::py_install("skimpy")
```

```{python}
# pip install pandas numpy
```

```{python}
#| code-fold: true
#| output: false

import os
import pandas as pd

def find_git_root(start=None):
    if start is None:
        start = os.getcwd()
    path = os.path.abspath(start)
    
    while True:
        if os.path.isdir(os.path.join(path, ".git")):
            return path
        
        parent = os.path.dirname(path)
        if parent == path:
            raise FileNotFoundError("No .git directory found — are you inside a Git repository?")
        
        path = parent

repo_root = find_git_root()
datasets_path = os.path.join(repo_root, "datasets")

# Reading the datafile credit_risk_dataset.csv
credit_path = os.path.join(datasets_path, "credit_risk_dataset.csv")
credit1 = pd.read_csv(credit_path)
```


```{python}
from skimpy import skim

skim(credit1)
```


## 4. Conclusion


## 5. References

::: {#refs}
:::