---
title: "Explainable Credit Risk Assessment using LightGBM"

author: "Supriya Sathe (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Demonstrate how machine learning and explainable AI can improve
credit risk prediction while maintaining transparency in financial
decision-making.

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## 1. Introduction  {.smaller}

The need for assessing credit risks is essential for banks to determine their potential defaulter customers and take decisions regarding loan issuance. While traditional models, such as logistic regression, are quite simple, they lack the ability to capture intricate patterns that may exist within vast amounts of financial data. However, while machine learning algorithms, such as LightGBM, have superior predictive capability, they tend to lack clarity since they are difficult to interpret.

## 2. Methods (LightGBM)

LightGBM Model assumes that the credit risk can be treated as a binary classification problem.

LightGBM Logic: This technique constructs trees progressively while minimizing the value of a loss function. The ultimate result is the combination of all the trees :

$$\hat{y}_i = \sum_{k=1}^K f_k(x_i)$$

SHAP Values: Based on Cooperative Game Theory, SHAP breaks down the prediction as follows:

$$f(x) = \phi_0 + \sum_{j=1}^M \phi_j$$


## 3. Analysis and Results

### 3.1 Data Preprocessing 
The dataset with 32,581 observations contains both financial, credit, and demographic attributes. The preprocessing process involves the following:

Handling Missing Values - Median replacement of variables like person_emp_length and loan_int_rate.
Categorical Variable Labeling - Labeling the categorical variables like loan_intent and person_home_ownership.
Separating Independent and Dependent Variables - X represents independent variables, whereas y represents dependent variable loan_status.
Training and Testing Data Split - Training data consists of 80% observations, and testing data includes 20% observations.

## 3.Analysis and Results

### **3.2. Dataset Visualization**

```{python}
#| label: corr-heatmap
#| fig-cap: "Correlation Heatmap of Numerical Features"

library(ggplot2)
library(reshape2)

# Select numeric columns
num_data <- credit_data %>% select(person_age, person_income, person_emp_length, loan_amnt, loan_int_rate, loan_%inc, cred_hist_len)

# Compute correlation matrix
corr_mat <- round(cor(num_data, use="complete.obs"), 2)
melted_corr <- melt(corr_mat)

# Plot heatmap
ggplot(melted_corr, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile(color="white") +
  scale_fill_gradient2(low="blue", high="red", mid="white", midpoint=0, limit=c(-1,1)) +
  theme_minimal() +
  labs(title="Correlation Heatmap of Numerical Features", x="", y="") +
  geom_text(aes(label=value), color="black", size=3) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.



## Data Exploration and Visualization {.smaller}



```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
